---
title: "ANTs longitudinal cortical thickness statistics: Region of interest version"
output: pdf_document
---

# Objective
Use regions of interest gained by multi-atlas labeling (MAL) to perform longitudinal statistics.

```{r fig.width=8, fig.height=4, echo=TRUE}
library(ANTsR)
library(plyr)
library(ggplot2)
library(date)
library(lme4)
library(nlme)
library(lmerTest)
library(packHV)
library(lattice)
library(visreg)
library(pheatmap)
library(effects)
##### define parameters #####
verbose<-1
basedir<-"/Users/stnava/data/LigerNinja/"
studytype<-"volume"
setwd(basedir)
data("DesikanKillianyTourville",package='ANTsR')
myRoiDescriptor<-DesikanKillianyTourville
woowoo<-1 # selects the weight value
WTVAL<-c(0,1)[woowoo]
wtsigma<-(-300.0)
outdir<-paste('/tmp/ADNIMXF_',studytype,round(WTVAL),sep='')
dir.create(outdir, showWarnings = FALSE)
p<-TRUE # do plots
##### parameters done #####
WTVAL<-as.numeric(WTVAL)
##############################
##############################
if ( ! exists("p") ) doggplot<-TRUE else doggplot<-FALSE
```

Define study demographic parameters.  
We use the demographics file to construct
the expected image file prefix.
```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE,error=FALSE}
### Most important thing - load a csv
### location of subject directories
demog<-read.csv(paste(basedir,'Data/adniCrossSectionalAntsSubset.csv',sep='/'))
ageAtBaseline<-rep(NA,nrow(demog))
deltaTime<-rep(NA,nrow(demog))
baseval1<-"Baseline"
baseval2<-"Screening"
for ( i in 1:nrow(demog) )
  {
  myid<-demog$ID[i]
  rawvisit<-as.character(demog$VISIT[i])
  numerictime<-strsplit( rawvisit, " " )[[1]]
  numerictime<-numerictime[ length(numerictime) ]
  if (  numerictime %in% c(baseval1,baseval2,"6","12","18","24","36") )
    {
    if ( numerictime == baseval1 | numerictime == baseval2 )
      {
      numerictime<-0
      } else numerictime<-as.numeric( numerictime )
    deltaTime[i]<-numerictime
    ageAtBaseline[i]<-min(  demog$AGE[ demog$ID == myid  ] )
    }
  }
if ( verbose > 1 )  print(sum(haveSubjectTimePoint))
demog<-cbind(demog,ageAtBaseline=ageAtBaseline,
  deltaTime=deltaTime)
```

Now find out how many subjects we have.
```{r countdata, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
# count the subjects
haveSubject<-!is.na(demog$ageAtBaseline)
subjectct<-sum( haveSubject )
print(paste("We have",subjectct,"subjects"))
```



Now compute the csv summary files from the ROI images.  
We compute volume, thickness and surface area for each
region that is available in the DKT labels that we
usually use.  The labels come from
[Arno's Mindboggle](http://mindboggle.info/data.html).

Let's quickly investigate these results.
```{r vizcorrs}
  myroidfThickness<-demog[, grep("thickness.",colnames(demog))]
  myroidfVolume<-demog[, grep("volume.",colnames(demog))]
  myroidfArea<-demog[, grep("area.",colnames(demog))]
  myroidfJ<-demog[, grep("jacobian.",colnames(demog))]
  meanthk<-rowMeans(myroidfThickness,na.rm=T)
  meanvol<-rowMeans(myroidfVolume,na.rm=T)
  meanare<-rowMeans(myroidfArea,na.rm=T)
  meanj<-rowMeans(myroidfJ,na.rm=T)
  print(cor.test(meanthk,meanj))
  print(cor.test(meanare,meanvol))
  if ( studytype == 'volume'  )
    myroidf<-myroidfVolume[haveSubject,]
  if ( studytype == 'area'  )
    myroidf<-myroidfArea[haveSubject,]
  if ( studytype == 'thickness'  )
    myroidf<-myroidfThickness[haveSubject,]
  if ( studytype == 'jacobian'  )
    myroidf<-myroidfJ[haveSubject,]
```

Run through the results with mixed effects modeling.

```{r mixeff, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
subdemog<-demog[haveSubject,c(1:21,355,356)] # just demographics
mydx<-as.factor( subdemog$DIAGNOSIS )
mydx = factor(mydx,levels(mydx)[c(4,3,2,1)])
rm("percentChangeByDX","pvalsByDX")
for ( jj in 1:ncol(myroidf) )
  {
  imging<-myroidf[,jj]
  if ( ! any(is.na(imging)) )
  {
  locdf<-data.frame( img=imging, SEX=subdemog$SEX,
    Age=subdemog$ageAtBaseline,
    MMSE=subdemog$MMSCORE,
    poi_dTime=subdemog$deltaTime,
    dT2=I(subdemog$deltaTime^2),
    poi_DX=(mydx),
    bvol=subdemog$brain.volume,
    sId=as.factor(subdemog$ID) )
  myform<-as.formula( img ~ bvol + Age + SEX + MMSE +
    poi_dTime*poi_DX  + (1+poi_dTime|sId) )
  myweights<-exp( (subdemog$deltaTime-WTVAL)^2 / wtsigma ) 
  myweights[ locdf$poi_DX == "Normal"  ]<-1
  myweights[is.na(myweights)]<-0
  myweights<-myweights/max(myweights)
  if ( jj == 1 )  hist(myweights)
  mdl<-lmer( myform , weights=myweights, data=(locdf))
  locallabel<-colnames(myroidf)[jj]
  smdl<-summary(mdl)$coeff
  if ( verbose == 1 ) {
    print(locallabel)
    print(smdl)  
    }
  if ( doggplot  ) {
#    plot( effect("poi_DX",mdl))
#    plot( effect("poi_dTime*poi_DX",mdl))
    locdf2=locdf
    rr<-residuals( lmer( img ~  (1|sId), data=locdf ))
    locdf2$img<-rr
    myform2<-as.formula( img ~ 1 + Age + SEX + Edu + poi_dTime*poi_DX)
    mdl2<-lm( myform2, weights=myweights , data=(locdf2))
    visreg(mdl2,"poi_dTime",by='poi_DX',main=locallabel)
    p<-ggplot(locdf, aes( x = poi_dTime , y = img, group = sId)) +  geom_line() + geom_smooth(aes(group = 1), method="lm", size = 2, se = T) + geom_point(aes(colour = factor(poi_DX))) + ggtitle(locallabel)
    plot(p)
  }
  if ( ncol(smdl) > 3 )
  {
  fxf<-fixef(mdl)
  baseval<-(fxf[1]+ mean(locdf$bvol)*fxf[grep("bvol",names(fxf))]
            + mean(locdf$Age)*fxf[grep("Age",names(fxf))]
            + 0.5 *fxf[grep("SEX",names(fxf))] )
  percentChange<-(baseval+fxf[grep("poi_",names(fxf))])/baseval
  if ( ! exists("percentChangeByDX") )
    {
    locnames<-names(fxf)[grep("poi_",names(fxf))]
    locnames<-sub(":","_",locnames)
    percentChangeByDX<-myroidf[1:length(percentChange),]*0
    rownames(percentChangeByDX)<-locnames
    }
  baserow<-(nrow(percentChangeByDX)-length(percentChange))+1
  locinds<-baserow:nrow(percentChangeByDX)
  percentChangeByDX[locinds,jj]<-(1.0-percentChange)*100
  if ( max(abs(percentChangeByDX[locinds,jj])) > 16000 ) derka
  poiInds<-grep("poi_",rownames(smdl))
  timemodelbase<-poiInds[1]
  if ( ! exists("pvalsByDX") )
    {
    locnames<-rownames(smdl)[poiInds]
    locnames<-sub(":","_",locnames)
    pvalsByDX<-myroidf[1:length(poiInds),]*0
    betasByDX<-myroidf[1:length(poiInds),]*0
    pvalsByDX[]<-NA
    rownames(pvalsByDX)<-locnames
    rownames(betasByDX)<-locnames
    }
  pvalsByDX[,jj]<-smdl[poiInds,5] # FIXME
  betasByDX[,jj]<-smdl[poiInds,1] # FIXME
  } # smdl ncol
  }
###############################################################
###############################################################
}
derka
```


Report some results
```{r reportRegions}
qvThresh<-0.01
myMeth<-"BH"
regionStats<-function( pvs, nm, pctchg ) {
  print(nm)
  mymin<-which.min(pvs)
  print(paste(colnames(myroidf)[mymin],
    p.adjust(pvs,myMeth)[mymin]))
  qv<-p.adjust(pvs,myMeth)
  if ( sum(qv <= qvThresh , na.rm=T ) >= 1 ) {
    cth<-max( pvs[ qv <= qvThresh ] , na.rm=T)
    print(paste("CriticalThresh",cth,"Rate",pctchg[mymin]))
    } else {
    cth<-min(pvs,na.rm=T)
    print(paste("Min-Uncorrected-P-val",cth,"Rate",pctchg[mymin]))
    }
  return(cth)
}
for ( i in 1:nrow(pvalsByDX) )
  {
  regionStats( as.numeric(pvalsByDX[i,]),
               rownames(pvalsByDX)[i],
               as.numeric(percentChangeByDX[i,]) )
  }
```


Set up visualization: see `antsSurfViz.sh`.
```{r vizsurf}
# write the images then do some system calls to antsSurf
opre<-paste(outdir,'/malf_viz_',studytype,sep='')
write.csv(percentChangeByDX,paste(opre,"Rates.csv",sep=''),row.names=F)
write.csv(pvalsByDX,paste(opre,"PVals.csv",sep=''),row.names=F)
```

## Heatmap of all region atrophy rates

```{r ratehm}
modperc<-percentChangeByDX[, ] # !is.na(colMeans(percentChangeByDX))
modpv<-pvalsByDX[,  ]# !is.na(colMeans(pvalsByDX))
qv<-data.matrix(modpv)
# maybe this is too conservative ... correcting the full matrix?
# qv<-matrix(p.adjust(data.matrix(modpv),method=myMeth),nrow=nrow(modpv))
for ( i in 1:nrow(qv) )
  qv[i,]<-p.adjust(as.numeric(modpv[i,]),method=myMeth)
colnames(qv)<-colnames(pvalsByDX)# [!is.na(colMeans(pvalsByDX))]
rownames(qv)<-rownames(pvalsByDX)
write.csv(qv,paste(opre,"QVals.csv",sep=''),row.names=F)
qrates<-percentChangeByDX# [,!is.na(colMeans(pvalsByDX))]
for ( k in 1:nrow(qv) ) {
  slct<-( qv[k,]<=qvThresh )
  slct[is.na(slct)]<-FALSE
  print(qrates[k,slct])
  }
```

```{r makevizimgs}
# for each reasonable region, map the rate to the malf image
if ( studytype=='MD') malf<-antsImageClone( eseg )
for ( k in 1:nrow(qv) ) {
  vizimg<-makevizimg( malf )
  bvizimg<-makevizimg( malf )
  slct<-( qv[k,]<=qvThresh )
  slct[is.na(slct)]<-FALSE
  names(slct)<-colnames(pvalsByDX)
  siginds<-which( slct )
  if ( length(siginds) > 0 )
  for ( j in 1:length(siginds) )
    {
    nm<-names(siginds)[j]
    labnum<-which(colnames(myroidf)==nm)
    labnum<-as.numeric( myRoiDescriptor$label_number[labnum] )
    lqrt<-qrates[k,siginds[j]]
    if ( lqrt < 0 & studytype != 'MD') lqrt<-0
    if ( studytype == 'MD' ) scl<-1.0 else scl<-(-1.0)
    lbeta<-betasByDX[k,siginds[j]]*scl
    vizimg[ malf == labnum ]<-lqrt*(scl*(-1.0))
    bvizimg[ malf == labnum ]<-lbeta
   # if ( verbose ) print(paste(lqrt,lbeta,1.0-qv[k,siginds[j]]))
    }
  onm<-paste(opre,rownames(qrates)[k],studytype,'Rate.nii.gz',sep='')
  antsImageWrite(vizimg,onm)
  onm<-paste(opre,rownames(betasByDX)[k],studytype,'Beta.nii.gz',sep='')
  antsImageWrite(bvizimg,onm)
  }
rownames(myroidf)<-demog$expectedImagePrefix[haveSubject]
rownames(myroidfThickness)<-demog$expectedImagePrefix[haveSubject]
write.csv(myroidf,paste(opre,studytype,'a.csv',sep=''))
write.csv(myroidfThickness,paste(opre,studytype,'b.csv',sep=''))
print(paste("TODO: Mean_Diffusion_Labels && TDP_3_Group"))
```
