---
title: "ANTs longitudinal cortical thickness statistics: Region of interest version"
output: pdf_document
---

# Objective
Use regions of interest gained by multi-atlas labeling (MAL) to perform longitudinal statistics.

```{r fig.width=8, fig.height=8, echo=TRUE}
library(ANTsR)
library(plyr)
library(ggplot2)
library(date)
library(lme4)
library(nlme)
library(lmerTest)
library(packHV)
library(lattice)
library(visreg)
library(pheatmap)
library(effects)
library(knitr)
##### define parameters #####
verbose<-1
basedir<-"/Users/stnava/data/LigerNinja/"
studytype<-"thickness"
processingstyle<-'long'
processingstyle<-'cross'
setwd(basedir)
data("DesikanKillianyTourville",package='ANTsR')
myRoiDescriptor<-DesikanKillianyTourville
WTVAL<-c( 0, 1, 2 )
wtsigma<-rep( 2.0, 3 )
outdir<-paste( '/tmp/ADNIMXF_', studytype, sep='')
dir.create( outdir, showWarnings = FALSE )
p<-TRUE # do plots
##### parameters done #####
##############################
##############################
if ( ! exists("p") ) doggplot<-TRUE else doggplot<-FALSE
doggplot<-FALSE
```

Define study demographic parameters.  
We use the demographics file to construct
the expected image file prefix.
```{r fig.width=8, fig.height=8, echo=TRUE,warnings=FALSE,message=FALSE,error=FALSE}
### Most important thing - load a csv
### location of subject directories
demogCross<-read.csv(paste(basedir,'Data/adniCrossSectionalAntsSubset.csv',sep='/'))
demogLong<-read.csv(paste(basedir,'Data/adniLongitudinalAntsSubset.csv',sep='/'))
if ( processingstyle == 'cross' ) demog<-demogCross
if ( processingstyle == 'long' ) demog<-demogLong

# sample the demog file for training
sdem<-rep( c(1,2,3,4), nrow(demog) )[1:nrow(demog)]
# demog<-demog[ sdem<3,   ]

# demog[474,]<-NA # FIXM
ageAtBaseline<-rep(NA,nrow(demog))
bvAtBaseline<-rep(NA,nrow(demog))
deltaTime<-rep(NA,nrow(demog))
baseval1<-"Baseline"
baseval2<-"Screening"
for ( i in 1:nrow(demog) )
  {
  myid<-demog$ID[i]
  rawvisit<-as.character(demog$VISIT[i])
  numerictime<-strsplit( rawvisit, " " )[[1]]
  numerictime<-numerictime[ length(numerictime) ]
  if (  numerictime %in% c(baseval1,baseval2,"6","12","18","24","36") )
    {
    if ( numerictime == baseval1 | numerictime == baseval2 )
      {
      numerictime<-0
      } else numerictime<-as.numeric( numerictime )
    deltaTime[i]<-numerictime
    ageAtBaseline[i]<-min(  demog$AGE[ demog$ID == myid  ] )
    deltaTime[i]<-demog$AGE[i]-ageAtBaseline[i]
    bvAtBaseline[i]<-max( demog$brain.volume[ demog$ID == myid  ] )
    }
  }
if ( verbose > 1 )  print(sum(haveSubjectTimePoint))
demog<-cbind(demog,ageAtBaseline=ageAtBaseline,
  deltaTime=deltaTime)
```

Now find out how many subjects we have.
```{r countdata, fig.width=8, fig.height=8, echo=TRUE,warnings=FALSE,message=FALSE}
# count the subjects
haveSubject<-!is.na(demog$ageAtBaseline)
subjectct<-sum( haveSubject )
print(paste("We have",subjectct,"subjects"))
```



Now compute the csv summary files from the ROI images.  
We compute volume, thickness and surface area for each
region that is available in the DKT labels that we
usually use.  The labels come from
[Arno's Mindboggle](http://mindboggle.info/data.html).

Let's quickly investigate these results.
```{r vizcorrs}
  myroidfThickness<-demog[, grep("thickness.",colnames(demog))]
  myroidfVolume<-demog[, grep("volume.",colnames(demog))]
  myroidfArea<-demog[, grep("area.",colnames(demog))]
  myroidfJ<-demog[, grep("jacobian.",colnames(demog))]
  meanthk<-rowMeans(myroidfThickness,na.rm=T)
  meanvol<-rowMeans(myroidfVolume,na.rm=T)
  meanare<-rowMeans(myroidfArea,na.rm=T)
  meanj<-rowMeans(myroidfJ,na.rm=T)
  print(cor.test(meanthk,meanj))
  print(cor.test(meanare,meanvol))
  if ( studytype == 'volume'  )
    myroidf<-myroidfVolume[haveSubject,]
  if ( studytype == 'area'  )
    myroidf<-myroidfArea[haveSubject,]
  if ( studytype == 'thickness'  )
    myroidf<-myroidfThickness[haveSubject,]
  if ( studytype == 'jacobian'  )
    myroidf<-myroidfJ[haveSubject,]
myroidf<-antsrimpute( myroidf )
```


```{r permset}
subdemog<-demog[haveSubject,c(1:21,355,356)] # just demographics
mydx<-as.factor( subdemog$DIAGNOSIS )
mydx = factor(mydx,levels(mydx)[c(4,3,2,1)])
mydx[ mydx == "LMCI" ]<-"MCI"
mydx = factor( mydx ) # make it just mci, ad, normal

# set up diagnosis data frame for each unique subject
uniqDX<-list()
for ( sid in unique( subdemog$ID ) )
  uniqDX[[ as.character(sid) ]]<-mydx[ subdemog$ID==sid ][1]
uniqDX<-data.frame( uid=unique( subdemog$ID ),
  path=unlist(uniqDX), pathperm=unlist(uniqDX)  )
nctl<-sum( uniqDX$path=="Normal", na.rm=T )
nmci<-sum( uniqDX$path=="MCI", na.rm=T )
nad<-sum( uniqDX$path=="AD", na.rm=T )
###################################
rm("percentChangeByDX","pvalsByDX")
# set up for permutation testing
nperms<-500+1
ndx<-2
roisrch<-grep( "thickness.", colnames(demog) )
names(roisrch)<-colnames(demog)[roisrch]
roisrch<-roisrch[ grep("thickness",names(roisrch)) ]
nroi<-length( roisrch )
permarrayD0<-rep( 0, nperms*ndx*nroi )
dim( permarrayD0 )<-c( ndx, nroi, nperms )
permarrayD1<-rep( 0, nperms*ndx*nroi )
dim( permarrayD1 )<-c( ndx, nroi, nperms )
permarrayD2<-rep( 0, nperms*ndx*nroi )
dim( permarrayD2 )<-c( ndx, nroi, nperms )
permarrayL1<-rep( 0, nperms*ndx*nroi )
dim( permarrayL1 )<-c( ndx, nroi, nperms )
permarrayL2<-rep( 0, nperms*ndx*nroi )
dim( permarrayL2 )<-c( ndx, nroi, nperms )
```

Run through the results with mixed effects modeling.

```{r mixeff, fig.width=8, fig.height=8, echo=TRUE,warnings=FALSE,message=FALSE}
# do tests
pvdf1<-data.frame(  region=names(roisrch),  
    D0=rep(NA, nroi ),
    D1=rep(NA, nroi ),
    D2=rep(NA, nroi ),
    L1=rep(NA, nroi ),
    L2=rep(NA, nroi ) )
pvdf2<-pvdf1
valdf1<-pvdf1
valdf2<-pvdf1
jj<-0
for ( roival in roisrch )
  {
  jj<-jj+1
  imging<-antsrimpute( myroidf[,roival] ) # FIXME check this
  if ( ! any(is.na(imging)) )
  {
  locdf<-data.frame( img=imging, Sex=subdemog$SEX,
    Age=subdemog$ageAtBaseline,
    MMSE=subdemog$MMSCORE,
    poi_dTime=( subdemog$deltaTime ),
    dT2=I(subdemog$deltaTime^2),
    kg=subdemog$WEIGHT_KG,
    poi_DX=(mydx),
    bvol=bvAtBaseline[haveSubject],
    sId=as.factor(subdemog$ID) )
  myform<-as.formula( img ~ 1 + Age + Sex + kg +
    poi_DX   + (1|sId) ) #  + (1+poi_dTime|sId)
  for ( lwtct in 1:length(WTVAL) )
    {
    myweights <-exp( -(subdemog$deltaTime-WTVAL[lwtct])^2 / wtsigma[lwtct] )
    myweights[ locdf$poi_DX == "Normal"  ]<-1
    myweights[is.na(myweights)]<-0
    myweights<-myweights/max(myweights)
    if ( jj == 1 )  hist(myweights)
    mdl<-lmer( myform , weights=myweights, data=(locdf) )
    locallabel<-names(roisrch)[jj]
    #
    smdl<-summary(mdl)$coeff
    if ( lwtct == 1 ) {
      permarrayD0[1,jj,1]   <- smdl[5,4]
      if ( nrow(smdl) > 5 ) permarrayD0[2,jj,1]   <- smdl[6,4]
      }
    if ( lwtct == 2 ) {
      permarrayD1[1,jj,1] <-   smdl[5,4]
      permarrayL1[1,jj,1] <- ( smdl[5,4] - permarrayD0[1,jj,1] ) * -1
      if ( nrow(smdl) > 5 ) {
        permarrayD1[2,jj,1]<-smdl[6,4]
        permarrayL1[2,jj,1]<-( smdl[6,4] - permarrayD0[2,jj,1] ) * -1
        }
      }
    if ( lwtct == 3 ) {
      permarrayD2[1,jj,1] <-   smdl[5,4]
      permarrayL2[1,jj,1] <- ( smdl[5,4] - permarrayD1[1,jj,1] ) * -1
      if ( nrow(smdl) > 5 ) {
        permarrayD2[2,jj,1]<-smdl[6,4]
        permarrayL2[2,jj,1]<-( smdl[6,4] - permarrayD1[2,jj,1] ) * -1
        }
      }
    print( smdl )
    }
    for ( perm in 2:nperms )
      {
      plocdf<-locdf
      permdx<-sample( uniqDX$path )
      for ( sid in unique( locdf$sId ) ) {
        selector <- plocdf$sId == sid
        plocdf$poi_DX[ selector ]<-permdx[ uniqDX$uid == sid ]
#        plocdf$poi_dTime[ selector ]<-sample(locdf$poi_dTime[selector])
        }
      lwtct<-0
      for ( lwt in WTVAL )
        {
        lwtct<-lwtct+1
        myweights <-exp( -(subdemog$deltaTime-WTVAL[lwtct])^2 / wtsigma[lwtct] )
        myweights[ locdf$poi_DX == "Normal"  ]<-1
        myweights[is.na(myweights)]<-0
        myweights<-myweights/max(myweights)
        pmdl<-lmer( myform , weights=myweights, data=(plocdf) )
        spmdl<-summary(pmdl)$coeff
        if ( ncol(spmdl) > 3 )
        {
        if ( lwtct == 1 ) {
          permarrayD0[1,jj,perm]<-spmdl[5,4]
          if ( nrow(smdl) > 5 ) permarrayD0[2,jj,perm]<-spmdl[6,4]
          }
        if ( lwtct == 2 ) {
          permarrayD1[1,jj,perm]<-spmdl[5,4]
          permarrayL1[1,jj,perm] <- ( spmdl[5,4] - permarrayD0[1,jj,perm] ) * -1
          if ( nrow(smdl) > 5 ) {
            permarrayD1[2,jj,perm]<-spmdl[6,4]
            permarrayL1[2,jj,perm]<-( spmdl[6,4] - permarrayD0[2,jj,perm] ) * -1
            }
          }
        if ( lwtct == 3 ) {
          permarrayD2[1,jj,perm]<-spmdl[5,4]
          permarrayL2[1,jj,perm] <- ( spmdl[5,4] - permarrayD1[1,jj,perm] ) * -1
          if ( nrow(smdl) > 5 ) {
            permarrayD2[2,jj,perm]<-spmdl[6,4]
            permarrayL2[2,jj,perm]<-( spmdl[6,4] - permarrayD1[2,jj,perm] ) * -1
            }
          }
        } # ncol test
      } # lwt perm weights
      } # nperms
    ## done with perms, now summarize
    dx1pD0<-sum( permarrayD0[1,jj,2:nperms] < permarrayD0[1,jj,1]  )/(nperms)
    dx1pD1<-sum( permarrayD1[1,jj,2:nperms] < permarrayD1[1,jj,1]  )/(nperms)
    dx1pD2<-sum( permarrayD2[1,jj,2:nperms] < permarrayD2[1,jj,1]  )/(nperms)
    dx1pL1<-sum( permarrayL1[1,jj,2:nperms] > permarrayL1[1,jj,1]  )/(nperms)
    dx1pL2<-sum( permarrayL2[1,jj,2:nperms] > permarrayL2[1,jj,1]  )/(nperms)
    dx2pD0<-sum( permarrayD0[2,jj,2:nperms] < permarrayD0[2,jj,1]  )/(nperms)
    dx2pD1<-sum( permarrayD1[2,jj,2:nperms] < permarrayD1[2,jj,1]  )/(nperms)
    dx2pD2<-sum( permarrayD2[2,jj,2:nperms] < permarrayD2[2,jj,1]  )/(nperms)
    dx2pL1<-sum( permarrayL1[2,jj,2:nperms] > permarrayL1[2,jj,1]  )/(nperms)
    dx2pL2<-sum( permarrayL2[2,jj,2:nperms] > permarrayL2[2,jj,1]  )/(nperms)
    p1a<-permarrayL1[1,jj,2:nperms]
    p1b<-permarrayL1[1,jj,1]
    h1<-hist( p1a*(-1), xlim=c(0,max(abs(c(p1b,p1a)))), col=rgb(1,0,0,0.5) )
    h2<-hist( rep(p1b*(-1),100),breaks=c(p1b*-0.99,p1b*-1.01),
              add=T, col=rgb(0,0,1,0.5) )
    if ( verbose == 1 ) {
      print( names(roisrch)[jj] )
      print( paste( "PERMD0: dx1pD0", dx1pD0, "dx2pD0", dx2pD0  )  )
      print( paste( "PERMD1: dx1pD1", dx1pD1, "dx2pD1", dx2pD1  )  )
      print( paste( "PERMD2: dx1pD2", dx1pD2, "dx2pD2", dx2pD2  )  )
      print( paste( "PERML1: dx1pL1", dx1pL1, "dx2pL1", dx2pL1  )  )
      print( paste( "PERML2: dx1pL2", dx1pL2, "dx2pL2", dx2pL2  )  )
      pvdf1[jj,2]<-dx1pD0
      pvdf1[jj,3]<-dx1pD1
      pvdf1[jj,4]<-dx1pD2
      pvdf1[jj,5]<-dx1pL1
      pvdf1[jj,6]<-dx1pL2
      pvdf2[jj,2]<-dx2pD0
      pvdf2[jj,3]<-dx2pD1
      pvdf2[jj,4]<-dx2pD2
      pvdf2[jj,5]<-dx2pL1
      pvdf2[jj,6]<-dx2pL2

      valdf1[jj,2]<-mean(permarrayD0[1,jj,2:nperms]) - permarrayD0[1,jj,1]
      valdf1[jj,3]<-mean(permarrayD1[1,jj,2:nperms]) - permarrayD1[1,jj,1]
      valdf1[jj,4]<-mean(permarrayD2[1,jj,2:nperms]) - permarrayD2[1,jj,1]
      valdf1[jj,5]<-(mean(permarrayL1[1,jj,2:nperms])-permarrayL1[1,jj,1])*(-1)
      valdf1[jj,6]<-(mean(permarrayL2[1,jj,2:nperms])-permarrayL2[1,jj,1])*(-1)

      valdf2[jj,2]<-mean(permarrayD0[2,jj,2:nperms]) - permarrayD0[2,jj,1]
      valdf2[jj,3]<-mean(permarrayD1[2,jj,2:nperms]) - permarrayD1[2,jj,1]
      valdf2[jj,4]<-mean(permarrayD2[2,jj,2:nperms]) - permarrayD2[2,jj,1]
      valdf2[jj,5]<-(mean(permarrayL1[2,jj,2:nperms])-permarrayL1[2,jj,1])*(-1)
      valdf2[jj,6]<-(mean(permarrayL2[2,jj,2:nperms])-permarrayL2[2,jj,1])*(-1)
      }
    } # na check
  } # roiloop
```

```{r writeout,echo=FALSE}
write.csv( pvdf1, paste(outdir,'/pvdf1.csv',sep='') )
write.csv( pvdf2, paste(outdir,'/pvdf2.csv',sep='') )
write.csv( valdf1, paste(outdir,'/valdf1.csv',sep='') )
write.csv( valdf2, paste(outdir,'/valdf2.csv',sep='') )
```

```{r visres,results='asis',echo=FALSE}
sigth<-0.1
ctarr<-data.frame( dx=c("mci","ad"),
  baseline=c(0,0),
  duration1=c(0,0),
  duration2=c(0,0),
  ddur1base=c(0,0),
  ddur2base=c(0,0)
  )
ind<-1
stageFunction<-function( dxl, b, x, ind=1, sigth=NA, k, opt=1 )
  {
  if ( opt == 1 ) {
      refval  <- x[dxl,k,  ind]
      permval <- x[dxl,k, -ind]
      return( abs(refval) - mean(abs(permval) ) + b )
    } else {
      refval  <- x[dxl,k,  ind]
      permval <- x[dxl,k, -ind]
#      return( sum( abs(refval) > abs(permval) ) + b )
      return( sum( refval > permval)  + b )
    }
  }
opter<-1
for ( dx in 1:2 ) {
  jj<-0
  for ( roival in roisrch )
    {
    jj<-jj+1
    ctarr[dx,2]<-stageFunction( dx, ctarr[dx,2], permarrayD0, k=jj, opt=opter )
    ctarr[dx,3]<-stageFunction( dx, ctarr[dx,3], permarrayD1, k=jj, opt=opter )
    ctarr[dx,4]<-stageFunction( dx, ctarr[dx,4], permarrayD2, k=jj, opt=opter )
    ctarr[dx,5]<-stageFunction( dx, ctarr[dx,5], permarrayL1, k=jj, opt=opter )
    ctarr[dx,6]<-stageFunction( dx, ctarr[dx,6], permarrayL2, k=jj, opt=opter )
    }
  }
ctarr2<-ctarr
ctarr2[,2:ncol(ctarr)]<-0
opter<-2
for ( dx in 1:2 ) {
  for (k in roisrch ) # count D0 values
    {
    ctarr2[dx,2]<-stageFunction( dx, ctarr2[dx,2], permarrayD0, k=jj, opt=opter )
    ctarr2[dx,3]<-stageFunction( dx, ctarr2[dx,3], permarrayD1, k=jj, opt=opter )
    ctarr2[dx,4]<-stageFunction( dx, ctarr2[dx,4], permarrayD2, k=jj, opt=opter )
    ctarr2[dx,5]<-stageFunction( dx, ctarr2[dx,5], permarrayL1*(-1), k=jj, opt=opter )
    ctarr2[dx,6]<-stageFunction( dx, ctarr2[dx,6], permarrayL2*(-1), k=jj, opt=opter )
    }
  }
kable( ctarr , caption = "Summary significance.", padding = 0, digits = 3 )
kable( ctarr2 , caption = "Summary significance.", padding = 0, digits = 3 )
```


Demographics
```{r, results='asis',echo=FALSE}
tq<-function(x,n=100){ round(x*n)/n }
tf<-function(x,f1=mean,f2=sd) {
paste( " ",tq( f1(x,na.rm=T)) , "/", tq( f2( x, na.rm=T ) ) )
}
ageonset<-subdemog$ageAtBaseline[ subdemog$deltaTime == 0 ]
summdemog<-data.frame(
  Group=c( "control","mci","ad") ,
  N_subs=c( sum(mydx=="Normal" & locdf$poi_dTime == 0,na.rm=T),
            sum(mydx=="MCI" & locdf$poi_dTime == 0,na.rm=T),
            sum(mydx=="AD" & locdf$poi_dTime == 0,na.rm=T)),
  N_imgs=c( sum(mydx=="Normal"), sum(mydx=="MCI"), sum(mydx=="AD")),
  Male= c(
    sum( locdf$Sex[mydx=="Normal" & locdf$poi_dTime == 0] == "M",na.rm=T ),
    sum( locdf$Sex[mydx=="MCI" & locdf$poi_dTime == 0] == "M",na.rm=T ),
    sum( locdf$Sex[mydx=="AD" & locdf$poi_dTime == 0] == "M",na.rm=T ) ),
  AgeAtBase=
    c(tf( locdf$Age[mydx=="Normal" & locdf$poi_dTime == 0] ),
      tf( locdf$Age[mydx=="MCI" & locdf$poi_dTime == 0] ),
      tf( locdf$Age[mydx=="AD" & locdf$poi_dTime == 0] ) ),
  MMSE=c(
    tf( subdemog$MMSCORE[mydx=="Normal" & locdf$poi_dTime == 0] ),
    tf( subdemog$MMSCORE[mydx=="MCI" & locdf$poi_dTime == 0] ),
    tf( subdemog$MMSCORE[mydx=="AD" & locdf$poi_dTime == 0] ) ),
  DeltaTime=c(
    tf( locdf$poi_dTime[mydx=="Normal"]  ),
    tf( locdf$poi_dTime[mydx=="MCI"]  ),
    tf( locdf$poi_dTime[mydx=="AD"]  ) ),
  MedFoll=c(
    median( table( locdf$sId[ mydx=="Normal" ] )[  table( locdf$sId[ mydx=="Normal" ] ) > 0]  ),
    median( table( locdf$sId[ mydx=="MCI" ] )[  table( locdf$sId[ mydx=="MCI" ] ) > 0]  ),
    median( table( locdf$sId[ mydx=="AD" ] )[  table( locdf$sId[ mydx=="AD" ] ) > 0]  ) )
#  CDGLOBAL=c( " .", " bvFTD=5", " bvFTD=7 / PPA=14" )
  )
# table(mydx[ mydx=="MCI" &  locdf$poi_dTime == 0 ])
# table(mydx[ mydx=="AD" &  locdf$poi_dTime == 0 ])
kable( summdemog , caption = "Subject demographics. MedFoll indicates median number of follow-up visits.", padding = 0, digits = 3 )
```


## Material and Procedures
